---
title: "01_Exploratory"
author: "Calen P. Ryan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
    
---
Originally written April 13th, 2020

# Setup

---

## Setup: Options 
```{r}
knitr::opts_chunk$set(tidy = TRUE, fig.width=10, fig.height=8) 
```


## Setup: Packages - messages hidden
```{r, warning = F, message = F}
library(tidyverse) # The tidyverse, baby
library(haven) # Used to load different datatypes, like .DTA or .dat
library(readr) # Also used to read different data types
library(lubridate) # Working with dates
library(expss) # 
library(sjlabelled) # as_factor, as_character
library(vcd)## used for estimating goodness of fit for poisson distribution
library(here)
```


# Load Data

```{r, warning = FALSE}
source(here("Code/survey_measures/01a_anthro.R"))
# source(here("Code/survey_measures/01b_physact.R"))
source(here("Code/survey_measures/01c_repro.R"))
source(here("Code/survey_measures/01d_health.R"))
source(here("Code/survey_measures/01e_testost.R"))
source(here("Code/survey_measures/01f_clocks.R"))
```

# Anthropometrics
```{r}
clock_anthro <-left_join(clocks, anthro, by = "uncchdid")


clock_anthro %>%
  select( bmi, whr, arm_musc_area, skinfold_sum, IEAA, EEAA, DNAmPhenoAgeAdjAge, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )

```

# Health
See also "testost" and "anthro" for measures of ic percent fat, kcal food consumed, other measures of fatness etc.
```{r}
clock_health <-left_join(clocks, health, by = "uncchdid")


clock_health %>%
  select(smoke, ofensmok, drink, oftendri, IEAA, EEAA, DNAmPhenoAgeAdjAge, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```

# Reproduction

```{r}
clock_repro <-left_join(clocks, repro, by = "uncchdid")



clock_repro %>%
  select(ageroman, agesexin, agecrush, agedate, agecourt, IEAA, EEAA, DNAmPhenoAgeAdjAge, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )

```
```{r}
clock_repro %>%
  select(romantic, sexinter, presrela, evermar, numbsex, new_numbpreg, IEAA, EEAA, DNAmPhenoAgeAdjAge, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```
```{r}
clock_repro %>%
  select(pairbond_score, sexactivity_score, risksex_score, IEAA, EEAA, DNAmPhenoAgeAdjAge, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```
```{r}
clock_repro %>%
  select(presrela, numbsex, CD8T:Gran, PlasmaBlast) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```

# Testosterone:

```{r}
clock_testost <-left_join(clocks, testost, by = "uncchdid")



clock_testost %>%
  select(amt05, pmt05, rightgrip_mean, leftgrip_mean, allele1_num, IEAA, EEAA, AgeAccelerationResidual, AgeAccelerationResidual, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )

```

Grip strength shows some interesting qualities. 
PMT and AMT do not
Men with > grip strength may have more physical work, develop/age faster. 

```{r}
clock_testost %>%
  select(rightgrip_mean, leftgrip_mean, fath05c, anyoffspring05, IEAA, EEAA, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```

```{r}
clock_testost %>%
  select(amt09, pmt09, fath09c, anyoffspring09, maristat09, IEAA, EEAA,  DNAmPhenoAgeAdjAge, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```
```{r}
clock_testost %>%
  select(amt05, pmt05, rightgrip_mean, leftgrip_mean, CD8T:Gran, PlasmaBlast) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```
```{r}
clock_testost %>%
  select(ickcal05, crowd05, icpcfat05, IEAA, EEAA, DNAmPhenoAgeAdjAge, DNAmGrimAgeAdjAge) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```

```{r}
clock_testost %>%
  select(ickcal05, crowd05, icpcfat05, CD8T:Gran, PlasmaBlast) %>%
  psych::pairs.panels(.,method = "pearson", # correlation method
                      hist.col = "gray80",
                      density = TRUE,  # show density plots
                      ellipses = TRUE,
                      stars = TRUE, # show correlation ellipses
                      cex.cor = 1
  )
```

