---
title: "Male_EWAS"
author: "Calen P. Ryan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
---

Originally written April 30th, 2020

# Setup: 

## Options 
```{r set options}
knitr::opts_chunk$set(tidy = TRUE, fig.width=10, fig.height=8) 
```

## Establishe here()
```{r establish here}
here::here()
```

## General Packages - messages hidden
```{r load general packages, warning = F, message = F}
library(tidyverse) # loads ggplot, tidyr, readr, dplyr, purrr, forcats, tibble, stringr
library(haven) # Used to load different datatypes, like .DTA or .dat
library(readr) # Also used to read different data types
library(sjlabelled) # as_factor, as_character
library(here)
library(knitr)
library(kableExtra)
```

## Bioconductor packages - messages hidden
```{r load bioconductor packages, warning = F, message = F}
library(limma)
library(Biobase)
```


# Load DNAm Data

## Select men with data
```{r}
men_w_eage <-read_csv(here::here("Data/Other", "men_w_eage.csv")) %>% 
  mutate(uncchdid = as_character(uncchdid))
```
## Get Mvals
```{r}
mvals <-data.table::fread(file = here::here("Data/DNAm", "mvals_cebu_final_Xincl.csv"),
                  select = c("V1", men_w_eage$uncchdid, "20727_R2")) 

mvals <-mvals %>% column_to_rownames("V1")

```
Don't worry bout the error - since 20727 is a replicate, the name doesn't match. I add it manually. 

## Filter to 5% variable for MEN
```{r, eval= FALSE}
# Loads the function for beta2m and m2beta
source(here::here("Code/EWAS", "EWAS_functions.R"))

# Originally used to calculate 5 percent variable. But loads as names_var_probes below. 
# source(here("Code/functions", "b-val_filter_5percent.R"))

```

# Five Percent probes for men only: 142,777 rows
```{r}
names_var_probes <-read_csv( here::here("Data/DNAm", "names_var_probes_menonly_xyincl.csv") )
```

## EWAS
### T Phenotype EWAS
```{r}
source(here::here("/Code/EWAS", "t_phenotype_ewas.R"))

amt_summary
pmt_summary
father_summary
hhkids_summary
icpcfat_summary
rgrip_summary
blood_t_summary

```

### Reproductive Effort EWAS
```{r }
source(here::here("/Code/EWAS", "repro_phenotype_ewas.R"))

romantic_summary
sexinter_summary
presrela_summary
numbsex_summary
new_numbpreg_summary
```

### Anthropometric EWAS
```{r anthro ewas}
source(here::here("/Code/EWAS", "anthro_phenotype_ewas.R"))

height_summary
bmi_summary
arm_musc_area_summary
skinfold_sum_summary
whr_summary
```


# Load Annotation 
```{r}
annotation <- read.csv(here::here("/Data/DNAm", "annotation_all.csv"), header = TRUE, nrows = 5e5) 

names(annotation)[1] <-"probe"

```


## Annotation for blood T
```{r}
blood_t_annot <-merge_annot(blood_t_toptable, annotation, pval = 0.1)
```
## Check out relationships
```{r}

dnam_t <-peep_dnam_pheno(top_annotation_file = blood_t_annot, pheno_file = testost)
 
dnam_t %>% 
  ggplot(., aes(x = log_blood_t , y = value, group = probe, color = probe))+
  geom_point(alpha = 0.3)+
  stat_smooth(geom = "line", alpha = 0.5)+
  ggpubr::theme_pubr()+
  theme(legend.position = "none")

dnam_t %>% 
  filter(probe %in% 
           c(blood_t_annot %>% 
           top_n(n = 14, wt = P.Value) %>% 
           pull(probe) )) %>% 
  ggplot(., aes(x = log_blood_t , y = value, group = probe, color = probe))+
  geom_point(alpha = 0.3)+
  stat_smooth(n = 4, geom = "line", alpha = 0.5)+
  ggpubr::theme_pubr()+
  theme(legend.position = "none")
```



### Annotation for numbsex and numbpreg
```{r}
numbsex_annot <-merge_annot(numbsex_toptable, annotation, pval = 0.1)
new_numbpreg_annot <-merge_annot(new_numbpreg_toptable, annotation, pval = 0.1)

```

```{r}
dnam_numbsex <-peep_dnam_pheno(top_annotation_file = numbsex_annot, pheno_file = repro)
dnam_new_numbpreg <-peep_dnam_pheno(top_annotation_file = new_numbpreg_annot, pheno_file = repro)

dnam_numbsex %>% 
  ggplot(., aes(x = numbsex , y = value, group = probe, color = probe))+
  geom_point(alpha = 0.3)+
  stat_smooth(geom = "line", alpha = 0.5)+
  ggpubr::theme_pubr()+
  theme(legend.position = "none")
  

dnam_new_numbpreg %>% 
  ggplot(., aes(x = new_numbpreg , y = value, group = probe, color = probe))+
  geom_point(alpha = 0.3)+
  stat_smooth(geom = "line", alpha = 0.5)+
  ggpubr::theme_pubr()+
  theme(legend.position = "none")
```



# All variables sites

## Five percent probes for men and women: 143,052 rows
```{r}
names_var_probes <-read_csv( here::here("Data/DNAm", "names_var_probes_m_and_f_xyincl.csv") )
```

# EWAS
## T Phenotype EWAS
```{r}
source(here::here("/Code/EWAS", "t_phenotype_ewas.R"))

amt_summary
pmt_summary
father_summary
hhkids_summary
icpcfat_summary
rgrip_summary
blood_t_summary

```

## Reproductive Effort EWAS
```{r repro ewas}
source(here::here("/Code/EWAS", "repro_phenotype_ewas.R"))

romantic_summary
sexinter_summary
presrela_summary
numbsex_summary
new_numbpreg_summary
```

## Anthropometric EWAS
```{r}
source(here::here("/Code/EWAS", "anthro_phenotype_ewas.R"))

height_summary
bmi_summary
arm_musc_area_summary
skinfold_sum_summary
whr_summary
```



