---
title: "Reproductive Status DNAm Analysis for EMPH"
author: "Calen P. Ryan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    theme: yeti
    toc: yes
    toc_depth: 5
editor_options: 
  chunk_output_type: inline
---
I added a correction that includes the correct pregord for parity to see if pregord tells us antying.
Many nulliparous women had pregord of 1 due to 2007 sample. This is corrected.




# Setup

---

## Load all methylation data and packages

Using source code and ff package for loading mvals data.frame
```{r Load data, results = "hide", echo = FALSE, warning = FALSE, message = FALSE, cached = TRUE}
library('readstata13')
library(tidyverse)
library("limma"); 
library("Biobase")
library(knitr)
library(kableExtra)
library(gplots)
library(here)

library(dendextend)
library(ggdendro)

# Load M-vals: ---- 

# FULL M-vals No correction
# source("/home/cpr183/Basic-Scripts/Load_mvals-NoCorrection.R", echo = TRUE)

# Full M-vals Corrected
source("/home/cpr183/Basic-Scripts/Load_mvals.R", echo = TRUE)


# Variable 5%
source("/home/cpr183/Basic-Scripts/Load_5%var_mvals_ProbesOnly.R", echo = TRUE)

```


## Load reproductive status data and labels and extract those in study:

### Load the repro.final
```{r}

repro <-read_csv("/home/cpr183/Parity_Pregnancy/Data/repro.final.csv")

labels <-readRDS("/home/cpr183/Parity_Pregnancy/Data/repro.final.labels.rds")

# Should you wish to look at the labelled variables. Quite a few are missing...
# str(labels, list.len=ncol(repro))

```

# Parity correct for nulliparous
```{r}
repro %>% 
  filter(reprostat == "nulliparous",!is.na(pregord))
# Some nulliparous women have pregord 1 in here (because pregord = 1 in 2007)

repro %>% 
  mutate(pregord = if_else(reprostat == "nulliparous", 
                           0, 
                           pregord)) %>% 
  group_by(reprostat, pregord) %>% 
  count() 
# Yes this looks right. 


repro <-repro %>% 
  mutate(pregord = if_else(reprostat == "nulliparous", 
                           0, 
                           pregord))
```

### Extract women with DNAm data
```{r}
# This should be a function!
ids <-substr(names(mvals)[-1], 2, 6)

repro <-repro %>% 
  filter(uncchdid %in% c(ids))

dim(repro)
# It doesn't change here as I used DNAm to figure out who I needed repro data for.
```

### Import genetic data

These are the first two principle components derived by Sarah Moore and used for McDade et al. 2019 AJPA.
```{r}

load("/home/cpr183/ch_assets/Data/CEBU_MDS2(1).RData")
# MDS2

genetic <-rownames_to_column(MDS2, var = "uncchdid"); rm(MDS2)

```


### Import Smoking status


```{r}
load("/home/cpr183/ch_assets/Data/CEBU_pheno_with_smoking.RData")
# pheno_extended

smoker <-pheno_extended %>% 
  select(uncchdid, smoker)

rm(pheno_extended)
```

### Import SES3 variable

This is the same SES variable used in Ryan et al. 2018 Scientific Reports. I did not calculate these myself and unfortunately still have not checked this yet. Talk to Dan and read our correspondances on this.
```{r warning=FALSE}

SES3 <-read.dta13(here("/Data/2017-12-01 file for Calen.dta"), missing.type= ,generate.factors = T) %>% 
  select(uncchdid, SES3, icage05_blood) 

```


## Merge to create repro.extra
```{r}
extra <-plyr::join_all(list(genetic, SES3, smoker), by='uncchdid', type='left')
rm(genetic, SES3, smoker)

repro$uncchdid <-as.character(repro$uncchdid)

repro.extra <-left_join(repro, extra, by = "uncchdid"); 
rm(extra)


repro.extra$reprostat <-as_factor(repro.extra$reprostat)

repro.extra$reprostat <-plyr::revalue(repro.extra$reprostat, c("nulliparous" = "Nulliparous",
                             "pregnant"=  "Pregnant", 
                             "breastfeeding" = "Breastfeeding", 
                             "preg_&_bf" = "Pregnant & Breastfeeding", 
                             "parous" = "Parous"))


repro.extra[repro.extra$reprostat == "Pregnant & Breastfeeding", "reprostat"] <-"Pregnant"

repro.extra1 <-repro.extra

repro.extra <-repro.extra %>% 
  select(uncchdid, reprostat, icage05_blood, MDS1, MDS2, SES3, smoker, pregord) %>%
  droplevels() %>% 
  na.omit() 
```

# CORRECTED Mvals - create matrix
```{r, results = "hide", echo = FALSE}

repro.extra$reprostat <-droplevels(repro.extra$reprostat)



include <-repro.extra %>% 
  select(uncchdid, reprostat, icage05_blood, MDS1, MDS2, SES3, smoker, pregord) %>% 
  na.omit() %>% 
  select(uncchdid)


# Search for the uncchdid columns within "no.data" object 
matches <-unique( grep(paste(include$uncchdid, collapse="|"),
                       names(mvals), value = TRUE))


# And extract
mvals.drop <-mvals[, c(1, which(names(mvals) %in% c(matches)))]



```

The new matrix has the following dimensions...`r dim(mvals.drop)`


```{r}
################
rowz <-nrow(mvals.drop)
colz <-ncol(mvals.drop[,-1])

matrix.corrected <-matrix(sapply(c(mvals.drop[,-1]), as.numeric), nrow= rowz, ncol=colz , byrow=FALSE)
# So it took at least 11 minutes to create this matrix. Faster than my computer? Maybe not. 

matrix.corrected[1:4,1:4]
############

rownames(matrix.corrected) <-mvals[,1]
colnames(matrix.corrected) <-names(mvals.drop[,-1])


# Are mvals for a given row and column consistent between datasets (mvals.uncor and mvals.drop)?
# Check to make sure:
#mvals[mvals$probe =="cg00000029", "X23233"]
#matrix.corrected["cg00000029", "X23233"]


mvals[mvals$probe =="cg03340466", "X23104"]
matrix.corrected["cg03340466", "X23104"]

#  Ok. I now have the matrix. What next? 
#  - Design matrix.
#  - Run limma
#  - What else, man? 
```

```{r}
# colnames(matrix)
# focals2$uncchdid
identical(substr(colnames(matrix.corrected),2,6), repro.extra$uncchdid)


# Reorder:
x <- substr(colnames(matrix.corrected),2,6)

repro.extra <-repro.extra %>%
  slice(match(x, uncchdid))


identical(substr(colnames(matrix.corrected),2,6), repro.extra$uncchdid)
# Good!

```



# PARITY: Run the CORRECTED EWAS
```{r}

repro.extra %>% 
  select(reprostat, pregord)

# Setup the design matrix. ---- 
design <- model.matrix(~icage05_blood + MDS1 + MDS2 + SES3 + smoker + reprostat * pregord, data = repro.extra);




dim(design)
dim(matrix.corrected)


# Run the EWAS and smooth the variance ---- 
repro.cor.ewas <-lmFit(matrix.corrected, design, na.action = na.omit)
#repro.cor.ewas <- contrasts.fit(repro.cor.ewas, contrast.matrix) 
repro.cor.ewas <- eBayes(repro.cor.ewas)
repro.cor.ewas.var5 <-repro.cor.ewas[c(mvals.var5[,1]), ]

# Pull only probes with >5% variability ---- 
repro.cor.ewas.var5.results <-decideTests(repro.cor.ewas.var5, adjust.method = "BH", p.value = 0.05)

# Summary data ---- 
summary(repro.cor.ewas.var5.results) %>%
  kable("html") %>%
  kable_styling()
# Ok, for all of them there are SOME significantly different probes...

```







# CORRECTED Mvals - create matrix
```{r, results = "hide", echo = FALSE}


repro.extra1 <-repro.extra %>% 
  filter(reprostat == "Parous" | reprostat == "Nulliparous")

repro.extra$reprostat1 <-droplevels(repro.extra$reprostat1)



include <-repro.extra1 %>% 
  select(uncchdid, reprostat, icage05_blood, MDS1, MDS2, SES3, smoker, pregord) %>% 
  na.omit() %>% 
  select(uncchdid)


# Search for the uncchdid columns within "no.data" object 
matches <-unique( grep(paste(include$uncchdid, collapse="|"),
                       names(mvals), value = TRUE))


# And extract
mvals.drop <-mvals[, c(1, which(names(mvals) %in% c(matches)))]



```

The new matrix has the following dimensions...`r dim(mvals.drop)`


```{r}
################
rowz <-nrow(mvals.drop)
colz <-ncol(mvals.drop[,-1])

matrix.corrected <-matrix(sapply(c(mvals.drop[,-1]), as.numeric), nrow= rowz, ncol=colz , byrow=FALSE)
# So it took at least 11 minutes to create this matrix. Faster than my computer? Maybe not. 

matrix.corrected[1:4,1:4]
############

rownames(matrix.corrected) <-mvals[,1]
colnames(matrix.corrected) <-names(mvals.drop[,-1])


# Are mvals for a given row and column consistent between datasets (mvals.uncor and mvals.drop)?
# Check to make sure:
#mvals[mvals$probe =="cg00000029", "X23233"]
#matrix.corrected["cg00000029", "X23233"]


mvals[mvals$probe =="cg03340466", "X23104"]
matrix.corrected["cg03340466", "X23104"]

#  Ok. I now have the matrix. What next? 
#  - Design matrix.
#  - Run limma
#  - What else, man? 
```

```{r}
# colnames(matrix)
# focals2$uncchdid
identical(substr(colnames(matrix.corrected),2,6), repro.extra1$uncchdid)


# Reorder:
x <- substr(colnames(matrix.corrected),2,6)

repro.extra <-repro.extra1 %>%
  slice(match(x, uncchdid))


identical(substr(colnames(matrix.corrected),2,6), repro.extra1$uncchdid)
# Good!

```



# PARITY: Run the CORRECTED EWAS
```{r}

repro.extra1 %>% 
  select(reprostat, pregord)

# Setup the design matrix. ---- 
design <- model.matrix(~icage05_blood + MDS1 + MDS2 + SES3 + smoker + as.factor(pregord), data = repro.extra1);




dim(design)
dim(matrix.corrected)


# Run the EWAS and smooth the variance ---- 
repro.cor.ewas <-lmFit(matrix.corrected, design, na.action = na.omit)
#repro.cor.ewas <- contrasts.fit(repro.cor.ewas, contrast.matrix) 
repro.cor.ewas <- eBayes(repro.cor.ewas)
repro.cor.ewas.var5 <-repro.cor.ewas[c(mvals.var5[,1]), ]

# Pull only probes with >5% variability ---- 
repro.cor.ewas.var5.results <-decideTests(repro.cor.ewas.var5, adjust.method = "BH", p.value = 0.05)

# Summary data ---- 
summary(repro.cor.ewas.var5.results) %>%
  kable("html") %>%
  kable_styling()
# Ok, for all of them there are SOME significantly different probes...

```



# CORRECTED Mvals - create matrix
```{r, results = "hide", echo = FALSE}


repro.extra2 <-repro.extra %>% 
  filter(reprostat == "Parous")

repro.extra$reprostat2 <-droplevels(repro.extra$reprostat2)



include <-repro.extra2 %>% 
  select(uncchdid, reprostat, icage05_blood, MDS1, MDS2, SES3, smoker, pregord) %>% 
  na.omit() %>% 
  select(uncchdid)


# Search for the uncchdid columns within "no.data" object 
matches <-unique( grep(paste(include$uncchdid, collapse="|"),
                       names(mvals), value = TRUE))


# And extract
mvals.drop <-mvals[, c(1, which(names(mvals) %in% c(matches)))]



```

The new matrix has the following dimensions...`r dim(mvals.drop)`


```{r}
################
rowz <-nrow(mvals.drop)
colz <-ncol(mvals.drop[,-1])

matrix.corrected <-matrix(sapply(c(mvals.drop[,-1]), as.numeric), nrow= rowz, ncol=colz , byrow=FALSE)
# So it took at least 11 minutes to create this matrix. Faster than my computer? Maybe not. 

matrix.corrected[1:4,1:4]
############

rownames(matrix.corrected) <-mvals[,1]
colnames(matrix.corrected) <-names(mvals.drop[,-1])


# Are mvals for a given row and column consistent between datasets (mvals.uncor and mvals.drop)?
# Check to make sure:
#mvals[mvals$probe =="cg00000029", "X23233"]
#matrix.corrected["cg00000029", "X23233"]


mvals[mvals$probe =="cg03340466", "X23104"]
matrix.corrected["cg03340466", "X23104"]

#  Ok. I now have the matrix. What next? 
#  - Design matrix.
#  - Run limma
#  - What else, man? 
```

```{r}
# colnames(matrix)
# focals2$uncchdid
identical(substr(colnames(matrix.corrected),2,6), repro.extra2$uncchdid)


# Reorder:
x <- substr(colnames(matrix.corrected),2,6)

repro.extra2 <-repro.extra2 %>%
  slice(match(x, uncchdid))


identical(substr(colnames(matrix.corrected),2,6), repro.extra2$uncchdid)
# Good!

```



# PARITY: Run the CORRECTED EWAS
```{r}

# Setup the design matrix. ---- 
design <- model.matrix(~0 + icage05_blood + MDS1 + MDS2 + SES3 + smoker + pregord, data = repro.extra2);




dim(design)
dim(matrix.corrected)


# Run the EWAS and smooth the variance ---- 
repro.cor.ewas <-lmFit(matrix.corrected, design, na.action = na.omit)
#repro.cor.ewas <- contrasts.fit(repro.cor.ewas, contrast.matrix) 
repro.cor.ewas <- eBayes(repro.cor.ewas)
repro.cor.ewas.var5 <-repro.cor.ewas[c(mvals.var5[,1]), ]

# Pull only probes with >5% variability ---- 
repro.cor.ewas.var5.results <-decideTests(repro.cor.ewas.var5, adjust.method = "BH", p.value = 0.05)

# Summary data ---- 
summary(repro.cor.ewas.var5.results) %>%
  kable("html") %>%
  kable_styling()
# Ok, for all of them there are SOME significantly different probes...

```

# Trimester/Gestation

```{r}
repro.weeks <-repro.extra1 %>% 
  filter(was.preg == "TRUE") %>% 
  filter(!is.na(preg.weeks))

```

```{r}
repro.weeks %>% 
  ggplot(., aes(x = preg.weeks))+
  geom_histogram()

repro.weeks %>% 
  filter(preg.weeks > 60)
```



# CORRECTED Mvals - create matrix
```{r, results = "hide", echo = FALSE}
include <-repro.weeks %>% 
  select(uncchdid, preg.weeks, icage05_blood, MDS1, MDS2, SES3, smoker, pregord) %>% 
  na.omit() %>% 
  select(uncchdid)


# Search for the uncchdid columns within "no.data" object 
matches <-unique( grep(paste(include$uncchdid, collapse="|"),
                       names(mvals), value = TRUE))


# And extract
mvals.drop <-mvals[, c(1, which(names(mvals) %in% c(matches)))]

```


```{r}
################
rowz <-nrow(mvals.drop)
colz <-ncol(mvals.drop[,-1])

matrix.corrected <-matrix(sapply(c(mvals.drop[,-1]), as.numeric), nrow= rowz, ncol=colz , byrow=FALSE)
# So it took at least 11 minutes to create this matrix. Faster than my computer? Maybe not. 

matrix.corrected[1:4,1:4]
############

rownames(matrix.corrected) <-mvals[,1]
colnames(matrix.corrected) <-names(mvals.drop[,-1])


# Are mvals for a given row and column consistent between datasets (mvals.uncor and mvals.drop)?
# Check to make sure:
#mvals[mvals$probe =="cg00000029", "X23233"]
#matrix.corrected["cg00000029", "X23233"]


mvals[mvals$probe =="cg03340466", "X22417"]
matrix.corrected["cg03340466", "X22417"]

#  Ok. I now have the matrix. What next? 
#  - Design matrix.
#  - Run limma
#  - What else, man? 
```


```{r}
# colnames(matrix)
# focals2$uncchdid
identical(substr(colnames(matrix.corrected),2,6), repro.weeks$uncchdid)


# Reorder:
x <- substr(colnames(matrix.corrected),2,6)

repro.weeks <-repro.weeks %>%
  slice(match(x, uncchdid))


identical(substr(colnames(matrix.corrected),2,6), repro.weeks$uncchdid)
# Good!

```
# PregWeeks: Run the CORRECTED EWAS
```{r}

# Setup the design matrix. ---- 
design <- model.matrix(~icage05_blood + MDS1 + MDS2 + SES3 + preg.weeks, data = repro.weeks);




dim(design)
dim(matrix.corrected)


# Run the EWAS and smooth the variance ---- 
repro.cor.ewas <-lmFit(matrix.corrected, design, na.action = na.omit)
#repro.cor.ewas <- contrasts.fit(repro.cor.ewas, contrast.matrix) 
repro.cor.ewas <- eBayes(repro.cor.ewas)
repro.cor.ewas.var5 <-repro.cor.ewas[c(mvals.var5[,1]), ]

# Pull only probes with >5% variability ---- 
repro.cor.ewas.var5.results <-decideTests(repro.cor.ewas.var5, adjust.method = "BH", p.value = 0.05)

# Summary data ---- 
summary(repro.cor.ewas.var5.results) %>%
  kable("html") %>%
  kable_styling()
# Ok, for all of them there are SOME significantly different probes...

```



# Breastfeeding

```{r}
repro.bf <-repro.extra1 %>% 
  filter(was.lact == "TRUE") %>% 
  filter(!is.na(days.bt.bld.bf))

```

```{r}
repro.bf %>% 
  ggplot(., aes(x = days.bt.bld.bf))+
  geom_histogram()
```



# CORRECTED Mvals - create matrix
```{r, results = "hide", echo = FALSE}
include <-repro.bf %>% 
  select(uncchdid, days.bt.bld.bf, icage05_blood, MDS1, MDS2, SES3, smoker, pregord) %>% 
  na.omit() %>% 
  select(uncchdid)


# Search for the uncchdid columns within "no.data" object 
matches <-unique( grep(paste(include$uncchdid, collapse="|"),
                       names(mvals), value = TRUE))


# And extract
mvals.drop <-mvals[, c(1, which(names(mvals) %in% c(matches)))]

```


```{r}
################
rowz <-nrow(mvals.drop)
colz <-ncol(mvals.drop[,-1])

matrix.corrected <-matrix(sapply(c(mvals.drop[,-1]), as.numeric), nrow= rowz, ncol=colz , byrow=FALSE)
# So it took at least 11 minutes to create this matrix. Faster than my computer? Maybe not. 

matrix.corrected[1:4,1:4]
############

rownames(matrix.corrected) <-mvals[,1]
colnames(matrix.corrected) <-names(mvals.drop[,-1])


# Are mvals for a given row and column consistent between datasets (mvals.uncor and mvals.drop)?
# Check to make sure:
#mvals[mvals$probe =="cg00000029", "X23233"]
#matrix.corrected["cg00000029", "X23233"]


mvals[mvals$probe =="cg03340466", "X23143"]
matrix.corrected["cg03340466", "X23143"]

#  Ok. I now have the matrix. What next? 
#  - Design matrix.
#  - Run limma
#  - What else, man? 
```


```{r}
# colnames(matrix)
# focals2$uncchdid
identical(substr(colnames(matrix.corrected),2,6), repro.bf$uncchdid)


# Reorder:
x <- substr(colnames(matrix.corrected),2,6)

repro.bf <-repro.bf %>%
  slice(match(x, uncchdid))


identical(substr(colnames(matrix.corrected),2,6), repro.bf$uncchdid)
# Good!

```
# BF.weeks: Run the CORRECTED EWAS
```{r}

repro.bf <-repro.bf %>% 
  mutate(bf.category=cut(days.bt.bld.bf, breaks=c(Inf, 60 , 30, -Inf), labels=c("high","middle","low"))) 

repro.bf %>% 
  group_by(bf.category) %>% 
  count()

# Setup the design matrix. ---- 
design <- model.matrix(~icage05_blood + MDS1 + MDS2 + SES3 + bf.category, data = repro.bf);




dim(design)
dim(matrix.corrected)


# Run the EWAS and smooth the variance ---- 
repro.cor.ewas <-lmFit(matrix.corrected, design, na.action = na.omit)
#repro.cor.ewas <- contrasts.fit(repro.cor.ewas, contrast.matrix) 
repro.cor.ewas <- eBayes(repro.cor.ewas)
repro.cor.ewas.var5 <-repro.cor.ewas[c(mvals.var5[,1]), ]

# Pull only probes with >5% variability ---- 
repro.cor.ewas.var5.results <-decideTests(repro.cor.ewas.var5, adjust.method = "BH", p.value = 0.05)

# Summary data ---- 
summary(repro.cor.ewas.var5.results) 
# Ok, for all of them there are SOME significantly different probes...

```


# SESSION INFO
```{r}
sessionInfo()
```